---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# geohabnet

This package is expanded upon [Xing et al (2021)](https://academic.oup.com/bioscience/article/70/9/744/5875255). It add capabilities to customize parameter values using functions and see results of cropland connectivity risk index in the form of plots. The goal of `geohabnet` is to enable users to visualize cropland connectivity risk index using their own parameter values. The risk analysis includes 3 maps -

1.  Mean cropland connectivity

2.  Difference

3.  Variance

Package currently support crops sourced from `geodata::monfredaCrops()` and `geodata::spamCrops()`. This analysis produces 3 maps - mean, variance and difference for the crop risk index. There are multiple ways in which functions can be used - generate final outcome and then the intermediate outcomes for more sophisticated use cases. Refer to vignettes. This values are propagated to other functions for performing operations such distance matrix calculation. The values are set in `parameters.yaml` and it can be accessed using `get_parameters()`. See the usage below.

## Installation

The source version of package can be installed from [GitHub](https://github.com/GarrettLab/CroplandConnectivity/tree/main/geohabnet) with:

```{r setup}
if (!require("devtools")) {
  install.packages("devtools")
}

devtools::install_github("GarrettLab/CroplandConnectivity", subdir = "geohabnet")
```

## geohabnet Example

```{r example}
library(geohabnet)

param_file <- geohabnet::get_parameters()
# now edit the file
geohabnet::set_parameters(new_params = param_file)
```

Run the analysis using -

```{r eval=FALSE}
geohabnet::senstivity_analysis()
```

`parameters.yaml` stores the parameter and its values. It can be accessed and set using `get_parameters()` and `set_parameters()` respectively. By default risk analysis is run on global index, for which scales are present in `global_scales()` .

Refer to help using ?*geohabnet::fun* or *help(geohabnet::fun)*

Refer to article [*Analyzing risk index using cropland connectivity*](https://garrettlab.github.io/CroplandConnectivity/articles/analysis.html) for more elaborate description and usages of functions in this package.
