---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# geohabnet

This package is expanded upon [Xing et al (2021)](https://academic.oup.com/bioscience/article/70/9/744/5875255). It add capabilities to customize parameter values using functions and see results of cropland connectivty risk index in the form of plots. The goal of geohabnet is to enable users to visualize cropland connectivity risk index using their own parameter values.

## Installation

You can install the development version of geohabnet from [GitHub](GarrettLab/CroplandConnectivity) with:

``` r
# install.packages("devtools")
devtools::install_github("GarrettLab/CroplandConnectivity", subdir = "geohabnet")
```

## geohabnet Example

This is a basic example which shows you how to run sensitivity analysis on pre-loaded parameters.

```{r example}
library(geohabnet)

param_file <- geohabnet::get_parameters()
# now edit the file
geohabnet::set_parameters(new_parameters_file = param_file)
geohabnet::senstivity_analysis()
```
