---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->

[![Github Pages](https://github.com/GarrettLab/HabitatConnectivity/actions/workflows/pages/pages-build-deployment/badge.svg?branch=main){width="250" height="25"}](https://github.com/GarrettLab/HabitatConnectivity/actions/workflows/pages/pages-build-deployment) [![CRAN status](https://www.r-pkg.org/badges/version/geohabnet){width="100" height="25"}](https://CRAN.R-project.org/package=geohabnet)

<!-- badges: end -->

# geohabnet

This package expands on [Xing et al (2021)](https://academic.oup.com/bioscience/article/70/9/744/5875255). It adds capabilities to customize parameter values using functions and shows the results of habitat connectivity risk index in the form of plots. The goal of `geohabnet` is to enable users to visualize a habitat connectivity risk index using their own parameter values. The risk analysis outputs 3 maps -

1.  Mean habitat connectivity (based on a habitat connectivity index defined by the user)

2.  Difference in habitat connectivity

3.  Variance in habitat connectivity

This package currently supports crop maps sourced from `geodata::monfredaCrops()` and `geodata::spamCrops()`. This analysis produces the 3 maps listed above. There are multiple ways in which functions can be used - generate the final outcome and then the intermediate outcomes for more sophisticated use cases. The vignettes provide several examples. The output values are propagated to other functions for performing operations such as distance matrix calculation. The values are set in `parameters.yaml` and it can be accessed using `get_parameters()`. See the usage below.

## Installation

Package can either be installed from CRAN:

```{r cran}
install.packages("geohabnet")
```

or the source version of package can be installed from [GitHub](https://github.com/GarrettLab/HabitatConnectivity/) with:

```{r github, eval=FALSE}
if (!require("devtools")) {
  install.packages("devtools")
}

devtools::install_github("GarrettLab/HabitatConnectivity", subdir = "geohabnet")
```

## geohabnet Example

```{r example, eval=FALSE}
library(geohabnet)

param_file <- geohabnet::get_parameters()
# now edit the file
geohabnet::set_parameters(new_params = param_file)
```

Run the analysis using -

```{r eval=FALSE}
geohabnet::sensitivity_analysis()
```

`parameters.yaml` stores the parameter and its values. It can be accessed and set using `get_parameters()` and `set_parameters()` respectively. By default risk analysis is run on global index, for which scales are present in `global_scales()` .

Refer to help using ?*geohabnet::fun* or *help(geohabnet::fun)*

Refer to article [*Analyzing risk index using cropland connectivity*](https://garrettlab.github.io/HabitatConnectivity/articles/analysis.html) for more elaborate description and usages of functions in this package.
